---
title: "__TITLE__"
author: "matricaria.suaveolens@gmail.com"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
bibliography:
  - bibliography.bib
  - tools.bib
  - grateful-refs.bib
---

```{r setup}
#| label: setup
#| echo: false
#| cache: false

knitr::opts_chunk$set(echo = TRUE, fig.path='figures/', cache = TRUE, fig.width = 10)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries}
#| label: libraries
#| message: false
#| cache: false
#| include: false

library(readr)
library(tibble)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(RColorBrewer)
library(vegan)
library(kfigr)
library(knitr)
library(grateful)
library(patchwork)
```

```{r source-core}
#| label: source-core
#| include: false

# Shared rank_stats()/top_taxa()/assign_palette()/taxonomy_barplot_data() -- see that file
# for what each does and why. Copied into scripts/ alongside this qmd; not part of the
# default project scaffold, since not every project needs a sanity check.
source("scripts/sanity_check_core.R")
```

```{r constants}
#| label: constants
GTDB_TAXRANKS    <- c('domain', 'phylum', 'class', 'order', 'family', 'genus', 'species')
ORDERED_TAXRANKS <- c('domain', 'kingdom', 'phylum', 'class', 'order', 'family', 'genus', 'species', 'strain', 'subspecies')

# metatdenovo filenames are prefixed by <assembly_name>.<orfcaller_name> (e.g.
# "megahit.prodigal"), so reruns with different settings don't overwrite each other's
# output. Set this to match whichever run's files are actually in data/ --
# `Sys.glob("data/*.counts.tsv.gz")` lists what's there if it's not obvious.
PREFIX <- "__ASSEMBLY_ORFCALLER__"
```

```{r read-data}
#| label: read-data

counts <- read_tsv(sprintf("data/%s.counts.tsv.gz", PREFIX), show_col_types = FALSE) %>%
  mutate(sample = as.character(sample))

# Taxonomy source varies by run -- prefer EUKulele when present (fixed
# domain/phylum/.../species columns, no parsing needed); fall back to Diamond taxonomy
# otherwise (rank column names there are whatever --diamond_taxonomy_ranks used, so check
# the file's actual header if this project only has that one).
eukulele_file <- Sys.glob(sprintf("data/%s.*_eukulele.tsv.gz", PREFIX))
diamond_file  <- Sys.glob(sprintf("data/%s.*.diamond.taxonomy.tsv.gz", PREFIX))

taxonomy <- if (length(eukulele_file) > 0) {
  read_tsv(eukulele_file[1], show_col_types = FALSE)
} else {
  read_tsv(diamond_file[1], show_col_types = FALSE)
}

tbl_taxonomic <- counts %>%
  left_join(taxonomy, by = "orf")

# kofamscan-uniq.tsv.gz is already reduced to one row per ORF (best hit), which is what
# makes a plain left_join safe here -- no risk of an ORF fanning out to multiple rows.
kofam <- read_tsv(sprintf("data/%s.kofamscan-uniq.tsv.gz", PREFIX), show_col_types = FALSE)

tbl_functional <- counts %>%
  left_join(kofam %>% select(orf, ko, ko_definition), by = "orf")

sample_metadata <- read_tsv("data/sample_table.tsv", show_col_types = FALSE)
```

# Version history

# Summary

# Introduction

# Materials and Methods

## R and packages

This analysis was run with `r R.version$version.string`. Versions of packages used can be found in @tbl-cite-packages.

```{r tbl-cite-packages}
#| label: tbl-cite-packages
#| cache: false
#| tbl-cap: Versions of R and packages used in this analysis.

cite_packages(output = "table", pkgs = "Session", out.dir = getwd()) %>%
  kable()
```

# Results

## Taxonomic composition

```{r taxonomy-barplot}
#| label: fig-taxonomy-barplot
#| fig-cap: Phylum-level community composition, top taxa above 1% of TPM; `Other` includes both below-threshold and unclassified ORFs.

top_phyla <- top_taxa(tbl_taxonomic, phylum, tpm, threshold = 0.01)
phylum_pal <- assign_palette(top_phyla, variable = "phylum")
barplot_tbl <- taxonomy_barplot_data(tbl_taxonomic, phylum, tpm, top_phyla)

ggplot(barplot_tbl, aes(x = sample, y = value, fill = .display)) +
  geom_col() +
  scale_fill_manual(values = phylum_pal, name = "Phylum") +
  labs(x = NULL, y = "TPM") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## NMDS: taxonomic and functional

NMDS is rank-based and robust to the choice of transformation, so no transformation is
applied here -- Bray-Curtis dissimilarity, computed twice: once on family-level TPM
(taxonomic composition, matching the barplot above) and once on KEGG-ortholog-level TPM
(functional composition, from KOfamScan's best-hit-per-ORF table). Both exclude
undefined/unmatched ORFs before summing (not folded into an "Other" bucket, unlike the
barplot above), and both colour by the same `treatment` palette so the combined figure
below can share a single legend.

```{r nmds}
#| label: fig-nmds
#| fig-cap: NMDS of family-level TPM (left) and KEGG-ortholog-level TPM (right), Bray-Curtis, no transformation.
#| fig-width: 14

treatment_pal <- assign_palette(sort(unique(sample_metadata$treatment)), variable = "treatment", palette = "Set1")

nmds_taxonomic_data <- tbl_taxonomic %>%
  filter(!is.na(family)) %>%
  group_by(sample, family) %>%
  summarise(value = sum(tpm), .groups = "drop") %>%
  pivot_wider(names_from = family, values_from = value, values_fill = 0)

nmds_taxonomic_matrix <- as.matrix(select(nmds_taxonomic_data, -sample))
rownames(nmds_taxonomic_matrix) <- nmds_taxonomic_data$sample

# trace = 0 silences the per-iteration console output; ord$tries hitting trymax (default
# 20) just means it used its full random-start budget, not that anything failed.
ord_taxonomic <- metaMDS(nmds_taxonomic_matrix, distance = "bray", autotransform = FALSE, trace = 0)

scores_taxonomic_tbl <- as.data.frame(ord_taxonomic$points) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

p_taxonomic <- ggplot(scores_taxonomic_tbl, aes(x = MDS1, y = MDS2, colour = treatment, shape = site)) +
  geom_point(size = 3) +
  scale_colour_manual(values = treatment_pal) +
  labs(title = "Taxonomic (family)", subtitle = sprintf("stress = %.3f", ord_taxonomic$stress))

nmds_functional_data <- tbl_functional %>%
  filter(!is.na(ko)) %>%
  group_by(sample, ko) %>%
  summarise(value = sum(tpm), .groups = "drop") %>%
  pivot_wider(names_from = ko, values_from = value, values_fill = 0)

nmds_functional_matrix <- as.matrix(select(nmds_functional_data, -sample))
rownames(nmds_functional_matrix) <- nmds_functional_data$sample

ord_functional <- metaMDS(nmds_functional_matrix, distance = "bray", autotransform = FALSE, trace = 0)

scores_functional_tbl <- as.data.frame(ord_functional$points) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

p_functional <- ggplot(scores_functional_tbl, aes(x = MDS1, y = MDS2, colour = treatment, shape = site)) +
  geom_point(size = 3) +
  scale_colour_manual(values = treatment_pal) +
  labs(title = "Functional (KEGG ortholog)", subtitle = sprintf("stress = %.3f", ord_functional$stress))

p_taxonomic + p_functional + plot_layout(guides = "collect")
```

# Discussion

# References

```{r}
# vim: sw=2
```
