---
title: "__TITLE__"
author: "matricaria.suaveolens@gmail.com"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
bibliography:
  - bibliography.bib
  - tools.bib
  - grateful-refs.bib
---

```{r setup}
#| label: setup
#| echo: false

knitr::opts_chunk$set(echo = TRUE, fig.path='figures/', fig.width = 10)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries}
#| label: libraries
#| message: false
#| include: false

library(readr)
library(tibble)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(RColorBrewer)
library(vegan)
library(kfigr)
library(knitr)
library(grateful)
library(patchwork)
```

```{r source-core}
#| label: source-core
#| include: false

# Shared rank_stats()/top_taxa()/assign_palette()/taxonomy_barplot_data() -- see that file
# for what each does and why. Copied into scripts/ alongside this qmd; not part of the
# default project scaffold, since not every project needs a sanity check.
source("scripts/sanity_check_core.R")
```

```{r constants}
#| label: constants
GTDB_TAXRANKS    <- c('domain', 'phylum', 'class', 'order', 'family', 'genus', 'species')
ORDERED_TAXRANKS <- c('domain', 'kingdom', 'phylum', 'class', 'order', 'family', 'genus', 'species', 'strain', 'subspecies')
```

```{r read-data}
#| label: read-data

# ampliseq doesn't hand you relative abundance directly (pre-3.0), so it's computed here
# from raw ASV counts, per sample, *before* the taxonomy join below -- this is what makes
# "keep the pre-removal relab for undefined taxa" work without extra bookkeeping: the
# denominator already includes every ASV, classified or not.
#
# NOTE: verify the header of ASV_table.tsv against your actual file (`head -2
# data/ASV_table.tsv`) before trusting this as-is -- some ampliseq export paths prefix a
# leading "# ..." banner line before the real header (QIIME/biom convention); if that's
# the case here, add `skip = 1` to the read_tsv() call below.
asv_counts <- read_tsv("data/ASV_table.tsv", show_col_types = FALSE)
names(asv_counts)[1] <- "ASV_ID"

# ASV_tax.<database>.tsv -- database name varies by run, hence the glob. Rank columns
# (Kingdom/Phylum/.../Genus) are DADA2's own assignTaxonomy() output names, capitalized,
# regardless of which reference database was used.
asv_tax <- read_tsv(Sys.glob("data/ASV_tax.*.tsv")[1], show_col_types = FALSE)

tbl <- asv_counts %>%
  pivot_longer(-ASV_ID, names_to = "sample", values_to = "count") %>%
  group_by(sample) %>%
  mutate(relab = count / sum(count)) %>%
  ungroup() %>%
  left_join(asv_tax %>% select(-sequence), by = "ASV_ID")

sample_metadata <- read_tsv("data/sample_table.tsv", show_col_types = FALSE)
```

# Version history

# Summary

# Introduction

# Materials and Methods

## R and packages

This analysis was run with `r R.version$version.string`. Versions of packages used can be found in @tbl-cite-packages.

```{r tbl-cite-packages}
#| label: tbl-cite-packages
#| tbl-cap: Versions of R and packages used in this analysis.

cite_packages(output = "table", pkgs = "Session", out.dir = getwd()) %>%
  kable()
```

# Results

## Taxonomic composition

```{r taxonomy-barplot}
#| label: fig-taxonomy-barplot
#| fig-cap: Phylum-level community composition, top taxa above 1% relative abundance; `Other` is below-threshold classified ASVs, `Unassigned` is ASVs with no phylum classification.

top_phyla <- top_taxa(tbl, Phylum, relab, threshold = 0.01)
phylum_pal <- assign_palette(top_phyla, variable = "Phylum")
barplot_tbl <- taxonomy_barplot_data(tbl, Phylum, relab, top_phyla)

ggplot(barplot_tbl, aes(x = sample, y = value, fill = .display)) +
  geom_col() +
  scale_fill_manual(values = phylum_pal, name = "Phylum") +
  labs(x = NULL, y = "Relative abundance") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## NMDS

NMDS is rank-based and robust to the choice of transformation, so no transformation is
applied here -- Bray-Curtis dissimilarity on family-level relative abundance, with
undefined (unclassified) ASVs excluded before summing (not folded into an "Other" bucket,
unlike the barplot above).

```{r nmds}
#| label: fig-nmds
#| fig-cap: NMDS of family-level relative abundance (Bray-Curtis, no transformation).

nmds_data <- tbl %>%
  filter(!is.na(Family)) %>%
  group_by(sample, Family) %>%
  summarise(value = sum(relab), .groups = "drop") %>%
  pivot_wider(names_from = Family, values_from = value, values_fill = 0)

nmds_matrix <- as.matrix(select(nmds_data, -sample))
rownames(nmds_matrix) <- nmds_data$sample

# trace = 0 silences the per-iteration console output; ord$tries hitting trymax (default
# 20) just means it used its full random-start budget, not that anything failed.
ord <- metaMDS(nmds_matrix, distance = "bray", autotransform = FALSE, trace = 0)

scores_tbl <- as.data.frame(ord$points) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

treatment_pal <- assign_palette(sort(unique(sample_metadata$treatment)), variable = "treatment", palette = "Set1")

ggplot(scores_tbl, aes(x = MDS1, y = MDS2, colour = treatment, shape = site)) +
  geom_point(size = 3) +
  scale_colour_manual(values = treatment_pal) +
  labs(title = sprintf("NMDS, stress = %.3f", ord$stress))
```

# Discussion

# References

```{r}
# vim: sw=2
```
